\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-06 papers/esop2010/main.tex
go

%external TODO : proofs for Hashtbl
%external TODO : proofs of system
%

@@

% TODO TODO
%Finally, we introduce two symbols
%\begin{eqnarray*}
%  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
%  |_{ε} &:& ∀ e. \efft{ e ε } -> \efft{ ε }.
%  !! & : & ∀αρ e. α~\refml_ρ -> \efft{ ρ e } -> α
%\end{eqnarray*}
%The first one denotes a combining operation for stores; the resulting store
%contains all the mappings of the second store, as well as all the mappings of
%the first one, as long as they are not present in the second one.
%union of the two arguments, in the sense that its domain is the union of its
%arguments.

@@

The presented programming language, along with the specifications, can, in
fact be used to prove programs. The weakest precondition calculus presented in
the next section can be used to obtain proof obligations which, when proved,
guarantee the correctness of the program with respect to its specification. On
the other hand, in its presented form, and because of the presence of
aliasing, one would constantly need equalities and inequalities between
reference variables to be able to reason about a program.

@@

The second form of aliasing is when we use the same name for {\em different}
memory locations. This can happen when we allow the user to call
$\mathit{ref}$ several times with the same region instantiation. A simple
possibility is to  suppress the $\mathit{ref}$ function entirely and to
replace it by a \letml\refml\ construct which combines name creation and
memory allocation. In this way, there can of course be only one reference per
name. On the other hand, because of the side condition $ρ\notin τ$, the
reference cannot escape the scope of its \letml\refml\ construct, and so, a
function cannot return a reference which has been created in its body. This
restriction could be leveraged a bit if one separates name creation and
allocation, while still maintaining that one can only allocate a single
reference in each region. This is how it is done in the correctness proof of
this restriction. The presented \letml\refml\ construct is of course a special
case.
%TODO now, all operations on maps can be statically resolved.

@@

let stores f0 m =
  forall x : Z, mem x m = true -> get x m = f0 x

@@

let memo [e] (f0 : int -> int) (x : int) = 
  {stores f0 !!table}
  if mem x !table then get x !table
  else
    let z = f0 x in
    table := set x z !table;
    z
  {r : r = f0 x /\ stores f0 !!table}
\end{who}
  \caption{The code along with the specification of the {\tt memo} function}
  \label{fig:memo}
\end{figure}

@@

\subsection{ymemo}

TODO %TODO
\begin{figure}[tpb]
\begin{who}
  let realizes = .. .
  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
    { forall (k : [[int ->{table} int ]]),
        { realizes f0 k} ff k { r : realizes f0 r} }
    let rec f (x : int) =
      { stores f0 !!table }
      if mem x !table then get x !table
      else
        let z = ff f x in
        table := set x z !table; z
      {r : r = f0 x /\ stores f0 !!table}
    in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specification}
  \label{fig:ymemo}
\end{figure}

@@

\begin{figure}[tbp]
\begin{center}
    \begin{tabular}{ | r | r | r | r |}
    \hline
    Program & Ocaml Code & Code + Spec in \who & Proofs \\ \hline
    Memo/ymemo & 15 loc & 26 loc & 90 lop  \\ \hline
    {\tt Array} & 69 loc & 170 loc  & 320 lop  \\ \hline
    {\tt Hashtbl} (incomplete) & & &  \\ \hline
    Koda-Ruskey~\cite{KanigFilliatre09wml} & 10 loc & 30 loc & $\sim$700 lop  \\ \hline
    \end{tabular}
\end{center}
  \caption{A summary of all programs proved so far using the \who\ tool}
  \label{fig:whoproofs}
\end{figure}

@@

The proof development concerning the {\tt Hashtbl} module is ongoing work. The
difficult part of this development is the model of the \ocaml\ hash table; In
\ocaml, when one overwrites a previously existing entry in the table, the old
value still exists, and is only {\em shadowed} by the new key/value pair. If
one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
if several bindings for a key exist, the key may be encountered several times,
during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
this is an association list instead of a simple mapping, where keys in front
of the list can shadow the same key further down in the list. 
% TODO Currently, only
%two functions of this module have been proved correct.

>>>>

%external TODO : proofs of system
%

@@

Finally, we introduce three symbols
\begin{eqnarray*}
  \oplus &:&∀ e1 e2 . \efft{ e1 } -> \efft{ e2 } -> \efft { e1  e2 }\\
  |_{ε} &:& ∀e. \efft{ e ε } -> \efft{ ε }\\
  {\tt !!} & : & ∀αρe. α~\refml_ρ -> \efft{ ρ e } -> α.
\end{eqnarray*}
The first one denotes a combining operation for stores; the resulting store
contains all the mappings of the second store, as well as all the mappings of
the first one, as long as they are not present in the second one. The second
operations is indexed by an effect and it restricts a store to a certain
effect. Finally, the double bang {\tt !!} is the logic equivalent to the
lookup operator {\tt !}. It is pure and it is always with respect to a certain
state.

@@

The presented programming language, along with the specifications, can, in
fact be used to prove programs. The weakest precondition calculus presented in
the next section can be used to obtain proof obligations which, when proved,
guarantee the correctness of the program with respect to its specification. On
the other hand, in its presented form, and because of the presence of
aliasing, one would constantly need equalities and disequalities between
reference variables to be able to reason about a program.

@@

The second form of aliasing is when we use the same name for {\em different}
memory locations. This can happen when we allow the user to call
$\mathit{ref}$ several times with the same region instantiation. A simple
possibility is to  suppress the $\mathit{ref}$ function entirely and to
replace it by a \letml\refml\ construct which combines name creation and
memory allocation. In this way, there can of course be only one reference per
name. On the other hand, because of the side condition $ρ\notin τ$, the
reference cannot escape the scope of its \letml\refml\ construct, and so, a
function cannot return a reference which has been created in its body. This
restriction could be leveraged a bit if one separates name creation and
allocation, while still maintaining that one can only allocate a single
reference in each region. This is how it is done in the correctness proof of
this restriction. The presented \letml\refml\ construct is of course a special
case.

In alias-free programs, the operations $\oplus$ and $|_ε$ on stores we
previously introduced can be executed {\em statically}, i.e. just by looking
at the domains of the stores. Also, we statically know which value (or which
named component) the accessor {\tt !!} will get from a store.

@@

let stores f0 m =
  forall x : int, mem x m = true -> get x m = f0 x

@@

let memo (f0 : int -> int) (x : int) = 
  {stores f0 !!table}
  if mem x !table then get x !table
  else
    let z = f0 x in
    table := set x z !table;
    z
  {r : r = f0 x /\ stores f0 !!table}
\end{who}
  \caption{The code along with the specification of the {\tt memo} function}
  \label{fig:memo}
\end{figure}

@@

\subsection{A more powerful memoization function}

\begin{figure}[tpb]
\begin{who}
  let realizes (f0 : int -> int) (f : [[ int ->{table} int ]]) =
    forall (x : int),
    { stores f0 !!table } f x { r : r = f0 x /\ stores f0 !!table }

  let ymemo (f0 : int -> int) 
            (ff : (int ->{table} int) ->{} int ->{table} int) = 
    { forall (k : [[int ->{table} int ]]),
        { realizes f0 k} ff k { r : realizes f0 r} }
    let rec f (x : int) =
      { stores f0 !!table }
      if mem x !table then get x !table
      else
        let z = ff f x in
        table := set x z !table; z
      {r : r = f0 x /\ stores f0 !!table}
    in f
  { rf : realizes f0 rf }
\end{who}
\caption{The {\tt ymemo function} and its specification}
  \label{fig:ymemo}
\end{figure}

Our second, more complicated example is the recursive version of {\tt memo},
dubbed {\tt ymemo}. We want to be able to define a recursive function whose
recursive calls are automatically memoized. The basic idea of {\tt ymemo} is
analogous to the fixpoint combinator in λ-calculus. As an example, here is how
one would implement a memoized Fibonacci function
using ymemo (ignoring the logic argument $f0$):
\begin{ocaml}
  let fib = 
    ymemo (fun g x -> if x <= 1 then x else g (x-1) + g (x-2))
\end{ocaml}
The anonymous function uses its first argument {\tt f} to do recursive calls.

{\tt ymemo} achieves this by defining a recursive function {\tt f} inside its
body (see Fig~\ref{fig:ymemo}). This function, when called with an integer
argument {\tt x} first checks if the table contains {\tt x} and if not, calls
{\tt ff} {\em with {\tt f} itself} and x as argument. As before, it stores the
obtained result in the table and returns it. Now, the return value of {\tt
ymemo} is precisely this recursive function {\tt f}. The only difference with
the equivalent \ocaml\ code is the usage of a logical argument {\tt f0} which
stands for the (mathematical) function to be defined.

Let's move on to the specification of {\tt ymemo}. We first define the {\tt
realizes} predicate between a pure function {\tt f0} and an impure one
modifying the table. This predicate states that {\tt f} {\em realizes}, i.e.
returns the same result as, {\tt f0} and additionally, if the table stored
valid key/value pairs, this is also the case after executing {\tt f}.
Actually, this is precisely the specification of {\tt memo}, so we could
easily prove {\tt realizes f0~(memo f0)} for any pure {\tt f0}. The definition
of the predicate uses the syntactic sugar for Hoare triples we introduced in
section~\ref{sec:genpic}.

The recursive function {\tt f} in the body of {\tt ymemo} has the same spec as
{\tt memo}, and indeed here it is trivial to prove the postcondition of {\tt
ymemo}, which states that its return value realizes {\tt f0}. Now, in order to
prove that the specification of {\tt f} is indeed correct, the only
information we need is that {\tt ff k} realizes {\tt f0} if {\tt k} is a
function which realizes {\tt f0}.

@@

\begin{figure}[tbp]
\begin{center}
    \begin{tabular}{ | r | r | r | r |}
    \hline
    Program & Ocaml Code & Code + Spec in \who & Proofs \\ \hline
    Memo/ymemo & 15 loc & 26 loc & 90 lop  \\ \hline
    {\tt Array} & 69 loc & 170 loc  & 320 lop  \\ \hline
    Koda-Ruskey~\cite{KanigFilliatre09wml} & 10 loc & 30 loc & $\sim$700 lop  \\ \hline
    \end{tabular}
\end{center}
  \caption{A summary of all programs proved so far using the \who\ tool}
  \label{fig:whoproofs}
\end{figure}

@@

%The proof development concerning the {\tt Hashtbl} module is ongoing work. The
%difficult part of this development is the model of the \ocaml\ hash table; In
%\ocaml, when one overwrites a previously existing entry in the table, the old
%value still exists, and is only {\em shadowed} by the new key/value pair. If
%one calls {\tt Hashtbl.remove} using this key, the old value reappears. Also,
%if several bindings for a key exist, the key may be encountered several times,
%during a call of {\tt Hashtbl.iter}, for example. A possible way of modeling
%this is an association list instead of a simple mapping, where keys in front
%of the list can shadow the same key further down in the list. 
% Currently, only
%two functions of this module have been proved correct.
